---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Cat Seitz"
date: "2/2/23"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData) 
library(ggbreak) 
library(patchwork)
```

### Exercise 1

```{r load_data}

staff <- read_csv("data/instructional-staff.csv", show_col_types = FALSE)

```

After pivoting the data, the data frame should be 55 rows. 

```{r pivot_df}

staff_long <- staff %>%
  pivot_longer(-faculty_type, names_to = "year") %>%
  mutate(value=as.numeric(value))
  

```

```{r plot_staff_long}

staff_long %>%
  ggplot(aes(x = year,
             y = value,
             group = faculty_type,
             color = faculty_type)) +
  geom_line()+
  labs(title="Percentage of staff hires by year",
       x="Year",
       y="Percentage of hires (%)")+
  guides(color = guide_legend(title="Faculty Type"))


```



### Exercise 2

In order to show that the proportion of part-time faculty hiring has increased compared to other instructional staff types, I would make a new data frame with the average of the four other staff types and then create the same graph. 

```{r create-new-df}

staff_df <- staff_long %>%
  subset(faculty_type!="Part-Time Faculty")

staff_df<- staff_df %>%
  group_by(year)%>%
  summarise(mean_pct=mean(value),
            faculty_type="other_types")

pt_faculty <- staff_long %>%
  subset(faculty_type=="Part-Time Faculty")

```

```{r new_plot}

ggplot() +
geom_line(data=pt_faculty,aes(x = year,
           y = value,
           group = faculty_type,
           color = faculty_type))+
geom_line(data=staff_df,aes(x = year,
           y = mean_pct,
           group = faculty_type,
           color = faculty_type))+
  labs(title="Percentage of staff hires by year",
       x="Year",
       y="Percentage of hires (%)")+
  guides(color = guide_legend(title="Faculty Type"))

```


### Exercise 3

The goal to improve the fisheries graph is to plot the data with a scale break since it seems like there are only a few countries that capture/farm way more fish than the rest. First, I'm going to filter the data for countries that have a total harvest of over 100,000 tons. After the filter, we now have 82 countries. 

```{r import_fish_data}

fisheries <- read_csv("data/fisheries.csv", show_col_types = FALSE)

```

```{r filter_data}

fisheries <- fisheries%>%
  filter(total>100000)

```

```{r plot}

g <- ggplot(data=fisheries, 
            aes(x=country, y=total))+ #, group=country, color=country
  geom_point()
g2<- g+ scale_y_break(c(2350000,81400000))
g2+theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  labs(title = "Total number of fish harvested per country")
```

Since there are still 82 countries that capture/farm over 100,000 fish, our x-axis will always be super jumbled if we keep all 82 countries in one graph. We could (1) raise the bar of how many fish a country needs to capture/farm to get on the graph or (2) break down the graph into a several graphs based on continent. Possibly, putting country on the y-axis would give us more room to see the country. I could also try plotting captured and farmed fish separately, but this wouldn't fix our x-axis problem. 

I'm going to filter again for at least 50,000 fish in each capture and farm categories to decrease the number of countries that qualify for the graph and see if it's low enough to make a better graph. After doing this, the number of farmed fishery locations decreased to 40 and the number of captured fishery location stayed at 82. Then, I increased the number of capture fish to 100,000. Now, I will create two separate graphs for captured and farmed fish. I also rotated the x-axis labels to make them readable. 

```{r filter_data_capture}

fish_capture <- fisheries%>%
  filter(capture>100000)

fish_farm <- fisheries%>%
  filter(aquaculture>50000)

```

```{r plot_farmed}

g <- ggplot(data=fish_farm, 
            aes(x=country, y=aquaculture))+ #, group=country, color=country
  geom_point()
g2<- g+ scale_y_break(c(17000000,67000000))
g2 +theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  labs(title = "Number of fish farmed per country")
```

```{r plot_captured}

g <- ggplot(data=fish_capture, 
            aes(x=country, y=capture))+ #, group=country, color=country
  geom_point()
g2<- g+ scale_y_break(c(7000000,18500000))
g2 +theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  labs(title = "Number of fish captured per country")
```




### Smokers in Whickham
### Exercise 1

```{r whickham}

data(Whickham)

```

These data come from an observational study because there was no manipulation introduced. The researchers observed individuals age and smoking habits and then observed if they were alive 20 years later. 


### Exercise 2

There are 1314 observations in this dataset with each observation representing a single subject. 


### Exercise 3

There are 3 variables in this dataset: Outcome, Smoker?, and Age at initial observation. All variables are integers.


```{r variable_type}

typeof(Whickham$outcome)
typeof(Whickham$smoker)
typeof(Whickham$age)

```

```{r variable_graph}

Whickham %>%
ggplot(aes(x = outcome)) +
  geom_bar()
Whickham %>%
ggplot(aes(x = smoker)) +
  geom_bar()
Whickham %>%
ggplot(aes(x = age)) +
  geom_bar()

```

### Exercise 4

I would expect smokers would be more likely than non-smokers to be dead. We will have to do something about age when examining the relationship between smoking and health outcome. 


### Exercise 5

```{r smoker_outcomes}

whick <- Whickham %>%
  count(smoker, outcome)

```

```{r smoker_outcome_graph}

ggplot(Whickham, aes(x=outcome, fill=smoker))+
  geom_bar(position="dodge")+
  labs(title = "Health Outcomes of Smokers vs. Non-smokers",
       x="Health Outcome",
       y="Number of individuals")
    
ggplot(Whickham, aes(x=smoker, fill=outcome))+
  geom_bar(position="dodge")+
  labs(title = "Health Outcomes of Smokers vs. Non-smokers",
       x="Smoking Status",
       y="Number of individuals")

```

The data indicate that more of the dead individuals were not smokers. 31.4% of non-smokers are now dead and 23.9% of smokers are now dead. This statistic makes it seem like individuals are more likely to die if they are non-smokers. Again, this statistic may look different when we account for age at the initial observation. 


### Exercise 6

```{r create-age-variable}

Whickham <- Whickham %>%
  mutate(age_cat = case_when(age <= 44 ~ "18-44",
                             age > 44 & age <= 64 ~ "45-64",
                             age > 64 ~ "65+"))

```


### Exercise 7

```{r count-table}

Whickham %>%
  count(smoker, age_cat, outcome)

```


3.5% of 18-44 year old non-smokers died, while 5.26% of 18-44 year old smokers died. 26.5% of 45-64 year old non-smokers died, while 32.4% of 45-64 year old smokers died. 85.5% of 65+ year old non-smokers died, while 88% of 65+ smokers died. These results clearly display that individuals are more likely to die if they smoke than if they don't smoke. Without taking into account the age of each individual, the number of dead individuals from the 65+ non-smoker group highly swing the percentage of dead non-smokers, but these individuals very likely died from other factors (potentially old age) because the outcome was collected when these individuals were at least 85 years old.


```{r new-visualization}

Whickham %>%
  ggplot(aes(x=smoker, fill=outcome))+
    geom_bar(position="dodge")+
    facet_wrap(~age_cat)+
  labs(title = "Health Outcomes of Smokers vs. Non-smokers by Age Group",
       x="Smoking Status",
       y="Number of individuals")

```



