---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Cat Seitz"
date: "2/2/23"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData) 
library(ggbreak) 
library(patchwork)
```

### Exercise 1

```{r load_data}

staff <- read_csv("data/instructional-staff.csv", show_col_types = FALSE)

```

After pivoting the data, the data frame should be 55 rows. 

```{r pivot_df}

staff_long <- staff %>%
  pivot_longer(-faculty_type, names_to = "year") %>%
  mutate(value=as.numeric(value))
  

```

```{r plot_staff_long}

staff_long %>%
  ggplot(aes(x = year,
             y = value,
             group = faculty_type,
             color = faculty_type)) +
  geom_line()+
  labs(title="Percentage of staff hires by year",
       x="Year",
       y="Percentage of hires (%)")+
  guides(color = guide_legend(title="Faculty Type"))


```



### Exercise 2

In order to show that the proportion of part-time faculty hiring has increased compared to other instructional staff types, I would make a new data frame with the average of the four other staff types and then create the same graph. 

```{r create-new-df}

staff_df <- staff_long %>%
  subset(faculty_type!="Part-Time Faculty")

staff_df<- staff_df %>%
  group_by(year)%>%
  summarise(mean_pct=mean(value),
            faculty_type="other_types")

pt_faculty <- staff_long %>%
  subset(faculty_type=="Part-Time Faculty")

```

```{r new_plot}

ggplot() +
geom_line(data=pt_faculty,aes(x = year,
           y = value,
           group = faculty_type,
           color = faculty_type))+
geom_line(data=staff_df,aes(x = year,
           y = mean_pct,
           group = faculty_type,
           color = faculty_type))+
  labs(title="Percentage of staff hires by year",
       x="Year",
       y="Percentage of hires (%)")+
  guides(color = guide_legend(title="Faculty Type"))

```


### Exercise 3

The goal to improve the fisheries graph is to plot the data with a scale break since it seems like there are only a few countries that capture/farm way more fish than the rest. First, I'm going to filter the data for countries that have a total harvest of over 100,000 tons. After the filter, we now have 82 countries. 

```{r import_fish_data}

fisheries <- read_csv("data/fisheries.csv", show_col_types = FALSE)

```

```{r filter_data}

fisheries <- fisheries%>%
  filter(total>100000)

```

```{r plot}

g <- ggplot(data=fisheries, 
            aes(x=country, y=total))+ #, group=country, color=country
  geom_point()
g2<- g+ scale_y_break(c(2350000,81400000))
g2
```


### Exercise 4





### Exercise 5




### Exercise 6






### Exercise 7













